---
title: "Getting Started with Spatial Mark-Resight Models"
author: "Speedgoat"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

CURRENTLY NOT RUNNING

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

# Introduction

The `spdgt.secr.host` package implements spatial mark-resight (SMR) models for analyzing capture-recapture data with spatial information. This vignette demonstrates how to:

1. Format your data for SMR analysis
2. Fit SMR models using different configurations
3. Interpret model results
4. Compare different model variants

## Installation

You can install the released version of smr from CRAN with:

```{r eval=FALSE}
install.packages("smr")
```

Or install the development version from GitHub with:

```{r eval=FALSE}
# install.packages("devtools")
devtools::install_github("username/smr")
```

## Required Packages

```{r setup}
library(spdgt.secr.host)
library(dplyr)
library(doParallel)
```

# Data Preparation

Let's create a simulated dataset to demonstrate the model. First, we'll set up our study area and sampling design:

```{r data-setup}
# Set random seed for reproducibility
set.seed(123)

# Study area parameters
xlim <- c(0, 10)
ylim <- c(0, 10)
buff <- 2

# Trap and camera locations
n_traps <- 20
n_cameras <- 25

# Generate random trap locations
X1 <- data.frame(
  x = runif(n_traps, xlim[1], xlim[2]),
  y = runif(n_traps, ylim[1], ylim[2])
)

# Generate random camera locations
X2 <- data.frame(
  x = runif(n_cameras, xlim[1], xlim[2]),
  y = runif(n_cameras, ylim[1], ylim[2])
)
```

Next, we'll set up population parameters and generate capture histories:

```{r population-setup}
# Population parameters
N <- 50  # True population size
n_marked <- 20  # Number of marked individuals
K1 <- 5  # Number of marking occasions
K2 <- 10  # Number of resighting occasions

# Movement and detection parameters
sigma_d <- 0.5  # Detection scale parameter
sigma_p <- 1.0  # Movement scale parameter
lam0_mark <- 0.1  # Baseline marking rate
lam0_sight <- 0.05  # Baseline sighting rate

# Generate activity centers
s <- data.frame(
  x = runif(N, xlim[1], xlim[2]),
  y = runif(N, ylim[1], ylim[2])
)

# Initialize capture history arrays
y_mark <- array(0, dim = c(n_marked, n_traps, K1))
y_sight_marked <- array(0, dim = c(n_marked, n_cameras, K2))
y_sight_unmarked <- array(0, dim = c(N - n_marked, n_cameras, K2))

# Simulate some captures (simplified for demonstration)
# In real applications, you would simulate based on detection probabilities
for(i in 1:n_marked) {
  y_mark[i,,] <- matrix(rbinom(n_traps*K1, 1, lam0_mark), n_traps, K1)
  y_sight_marked[i,,] <- matrix(rbinom(n_cameras*K2, 1, lam0_sight), n_cameras, K2)
}
for(i in 1:(N-n_marked)) {
  y_sight_unmarked[i,,] <- matrix(rbinom(n_cameras*K2, 1, lam0_sight), n_cameras, K2)
}
```

We'll also generate some genetic data and telemetry locations:

```{r genetic-telemetry}
# Generate genetic data
n_loci <- 3
n_alleles <- 4

# Function to generate random genotypes
generate_genotype <- function(n_loci, n_alleles) {
  matrix(sample(1:n_alleles, n_loci, replace = TRUE), nrow = 1)
}

# Generate genetic data
G_marked <- do.call(rbind, replicate(n_marked, 
                                    generate_genotype(n_loci, n_alleles), 
                                    simplify = FALSE))
G_unmarked <- do.call(rbind, replicate(N - n_marked, 
                                      generate_genotype(n_loci, n_alleles), 
                                      simplify = FALSE))

# Create telemetry data
n_telemetry_locs <- 5
locs <- array(NA, dim = c(n_marked, n_telemetry_locs, 2))

# Generate telemetry locations for first 10 marked individuals
for(i in 1:10) {
  for(j in 1:n_telemetry_locs) {
    locs[i,j,] <- c(
      rnorm(1, s$x[i], sigma_p),
      rnorm(1, s$y[i], sigma_p)
    )
  }
}
```

# Model Fitting

## Data Organization

First, we need to organize our data into the format expected by the SMR functions:

```{r data-organization}
data <- list(
  y_mark = y_mark,
  y_sight_marked = y_sight_marked,
  y_sight_unmarked = y_sight_unmarked,
  X1 = as.matrix(X1),
  X2 = as.matrix(X2),
  K1 = K1,
  K2 = K2,
  buff = buff,
  n_marked = n_marked,
  G_marked = G_marked,
  G_unmarked = G_unmarked,
  IDlist = list(
    ncat = n_loci,
    nallele = rep(n_alleles, n_loci),
    IDcovs = replicate(n_loci, 1:n_alleles, simplify = FALSE)
  ),
  locs = locs
)
```

## Model Configuration

We'll set up the model input parameters:

```{r model-config}
input <- list(
  niter = 1000,  # Number of MCMC iterations
  nburn = 500,   # Number of burn-in iterations
  nthin = 2,     # Thinning interval
  M = 100,       # Data augmentation parameter
  model_choices = c(1,2,3,4),  # Run all model variants
  inits = list(
    lam0_mark = 0.05,
    lam0_sight = 0.01,
    sigma_d = 0.5,
    sigma_p = 1.0,
    psi = 0.5,
    gamma = replicate(n_loci, rep(1/n_alleles, n_alleles), simplify = FALSE)
  ),
  obstype = c("bernoulli", "poisson"),
  proppars = list(
    lam0_mark = 0.02,
    lam0_sight = 0.02,
    sigma_d = 0.1,
    sigma_p = 0.2,
    s1 = 0.3,
    s2 = 0.3,
    s2t = 0.1
  ),
  max_proppars <- list(lam0_mark = 100, lam0_sight = 100, sigma_d = 100,
                       sigma_p = 100, s1 = 100, s2 = 100, s2t = 100),
  min_proppars <- list(lam0_mark = .001, lam0_sight = .001, sigma_d = .001,
                       sigma_p = .001, s1 = .001, s2 = .001, s2t = .001),
  nswap = NA,
  storeLatent = TRUE, 
  storeGamma = TRUE, 
  IDup = "Gibbs"
)
```

## Running the Model

Now we can fit the model using the wrapper function:

```{r model-fitting, eval=FALSE}
results <- SMR_wrapper(data, input)
```

```{r show-results, eval=FALSE}
# View summary of results
results %>%
  spread(model, Mean) %>%
  arrange(Parameter)
```

# Understanding Model Output

The SMR model provides estimates for several key parameters:

1. Detection Parameters:
   - `lam0_mark`: Baseline probability of marking
   - `lam0_sight`: Baseline probability of resighting
   - `sigma_d`: Scale parameter for detection function

2. Movement Parameters:
   - `sigma_p`: Scale parameter for movement between marking and resighting

3. Population Parameters:
   - `N`: Population size estimate
   - `psi`: Data augmentation parameter
   - `D`: Density estimate (derived)

## Model Variants

The wrapper function can fit four different model variants:

1. "Tele and Mobile": 
   - Uses telemetry data
   - Allows activity centers to move
   - Best when you have telemetry data and expect movement

2. "Tele and No Mobile":
   - Uses telemetry data
   - Assumes static activity centers
   - Suitable for short study periods with telemetry

3. "No Tele and Mobile":
   - No telemetry data
   - Allows activity center movement
   - Use when you expect movement but lack telemetry

4. "No Tele and No Mobile":
   - Basic model
   - No telemetry, static centers
   - Most parsimonious when movement is minimal

# Advanced Features

## Handling Trap Operation

You can account for varying trap effort using the `tf1` and `tf2` parameters:

```{r trap-operation, eval=FALSE}
# Example of varying trap effort
tf1 <- matrix(rpois(n_traps*K1, 5), n_traps, K1)  # Trap nights for marking
tf2 <- matrix(rpois(n_cameras*K2, 5), n_cameras, K2)  # Camera nights for resighting

data$tf1 <- tf1
data$tf2 <- tf2
```

## Using Custom State Space

Instead of a buffer, you can define a custom state space using vertices:

```{r state-space, eval=FALSE}
# Example of custom state space
vertices <- data.frame(
  x = c(0, 10, 10, 0),
  y = c(0, 0, 10, 10)
)

data$vertices <- as.matrix(vertices)
data$buff <- NULL  # Remove buffer when using vertices
```

# Example 2: Camera-Only Data

In many studies, researchers may only have camera trap data without a physical capture component. Here's how to analyze such data:

```{r camera-only-setup}
# Set random seed for reproducibility
set.seed(456)

# Study area parameters
xlim <- c(0, 10)
ylim <- c(0, 10)
buff <- 2

# Camera locations only (no traps)
n_cameras <- 40  # More cameras since it's the only data source

# Generate random camera locations
X2 <- data.frame(
  x = runif(n_cameras, xlim[1], xlim[2]),
  y = runif(n_cameras, ylim[1], ylim[2])
)

# Population parameters
N <- 50  # True population size
n_marked <- 15  # Number of marked individuals (marked through other means)
K2 <- 20  # Number of resighting occasions (longer survey)

# Detection parameters
sigma_d <- 0.5  # Detection scale parameter
lam0_sight <- 0.08  # Baseline sighting rate (higher since no physical capture)

# Generate activity centers
s <- data.frame(
  x = runif(N, xlim[1], xlim[2]),
  y = runif(N, ylim[1], ylim[2])
)

# Initialize sighting arrays (no marking array needed)
y_sight_marked <- array(0, dim = c(n_marked, n_cameras, K2))
y_sight_unmarked <- array(0, dim = c(N - n_marked, n_cameras, K2))

# Simulate sightings
for(i in 1:n_marked) {
  y_sight_marked[i,,] <- matrix(rbinom(n_cameras*K2, 1, lam0_sight), n_cameras, K2)
}
for(i in 1:(N-n_marked)) {
  y_sight_unmarked[i,,] <- matrix(rbinom(n_cameras*K2, 1, lam0_sight), n_cameras, K2)
}

# Generate genetic data
n_loci <- 3
n_alleles <- 4

G_marked <- do.call(rbind, replicate(n_marked, 
                                    generate_genotype(n_loci, n_alleles), 
                                    simplify = FALSE))
G_unmarked <- do.call(rbind, replicate(N - n_marked, 
                                      generate_genotype(n_loci, n_alleles), 
                                      simplify = FALSE))

# Create telemetry data (if available)
n_telemetry_locs <- 5
locs <- array(NA, dim = c(n_marked, n_telemetry_locs, 2))

# Generate telemetry locations for first 8 marked individuals
for(i in 1:8) {
  for(j in 1:n_telemetry_locs) {
    locs[i,j,] <- c(
      rnorm(1, s$x[i], sigma_p),
      rnorm(1, s$y[i], sigma_p)
    )
  }
}

# Organize data for camera-only analysis
data_camera <- list(
  y_mark = NULL,  # No marking data
  y_sight_marked = y_sight_marked,
  y_sight_unmarked = y_sight_unmarked,
  X1 = NULL,  # No trap locations
  X2 = as.matrix(X2),
  K1 = 0,  # No marking occasions
  K2 = K2,
  buff = buff,
  n_marked = n_marked,
  G_marked = G_marked,
  G_unmarked = G_unmarked,
  IDlist = list(
    ncat = n_loci,
    nallele = rep(n_alleles, n_loci),
    IDcovs = replicate(n_loci, 1:n_alleles, simplify = FALSE)
  ),
  locs = locs
)

# Model configuration for camera-only data
input_camera <- list(
  niter = 1000,
  nburn = 500,
  nthin = 2,
  M = 100,
  mobile_center = FALSE,  # Simplified movement model for camera-only data
  model_choices = c(2,4),  # Only run models without movement
  inits = list(
    lam0_sight = 0.05,
    sigma_d = 0.5,
    psi = 0.5,
    gamma = replicate(n_loci, rep(1/n_alleles, n_alleles), simplify = FALSE)
  ),
  obstype = c("bernoulli", "bernoulli"),
  proppars = list(
    lam0_sight = 0.02,
    sigma_d = 0.1,
    s2 = 0.3,
    s2t = 0.1
  ),
  max_proppars <- list(lam0_mark = 100, lam0_sight = 100, sigma_d = 100,
                       sigma_p = 100, s1 = 100, s2 = 100, s2t = 100),
  min_proppars <- list(lam0_mark = .001, lam0_sight = .001, sigma_d = .001,
                       sigma_p = .001, s1 = .001, s2 = .001, s2t = .001),
  nswap = NA,
  storeLatent = TRUE, 
  storeGamma = TRUE, 
  IDup = "Gibbs"
)

```

This example shows several key differences from the full mark-resight scenario:

1. Data Structure:
   - No physical capture/marking data (`y_mark = NULL`)
   - No trap locations (`X1 = NULL`)
   - Typically more cameras and longer survey duration
   
2. Model Configuration:
   - Simpler movement model (often `mobile_center = FALSE`)
   - Fewer model variants to consider
   - Different initialization values
   
3. Inference:
   - Focus on spatial distribution and density
   - Less information about movement
   - Identification through cameras and genetics only

To fit the camera-only model:

```{r camera-model-fitting, eval=FALSE}
results_camera <- SMR_wrapper(data_camera, input_camera)

# View summary of results
results_camera %>%
  spread(model, Mean) %>%
  arrange(Parameter)
```

Key considerations for camera-only studies:

1. Camera Placement:
   - Need good spatial coverage
   - Consider habitat features and animal movement patterns
   - More cameras often needed than in combined designs

2. Survey Duration:
   - Typically longer than combined designs
   - Need sufficient resightings of marked individuals
   - Balance between duration and closure assumptions

3. Marking Process:
   - Animals must be marked through other means
   - Consider collaborative efforts with other projects
   - Ensure marking is representative of population

# Best Practices

1. **MCMC Settings**:
   - Run longer chains (increase `niter`) for complex models
   - Use multiple chains to assess convergence
   - Adjust `proppars` if acceptance rates are poor

2. **Model Selection**:
   - Start with simpler models
   - Add complexity (movement, telemetry) only if supported by data
   - Compare models using DIC or WAIC (when implemented)

3. **Data Quality**:
   - Ensure sufficient spatial coverage of traps/cameras
   - Check for spatial closure assumptions
   - Validate genetic identifications

# References

- Royle, J. A., & Young, K. V. (2008). A hierarchical model for spatial capture-recapture data. Ecology, 89(8), 2281-2289.
- Sollmann, R., Gardner, B., & Belant, J. L. (2012). How does spatial study design influence density estimates from spatial capture-recapture models? PLoS One, 7(4), e34575.
